---
title: "Spatial pattern analyses of disease intensity data"
author: "Daniel Heck, Kaique Alves, and Sarah Pethybridge"
date: "1/18/2022"
output: 
  html_document:
    number_sections: true 
    df_print: paged
    theme: cosmo
    toc: true
    toc_float:
      collapsed: yes
    self_contained: yes
---

# Packages

```{r warning=F}
# install.packages("pacman")  # Install the package in your computer
library(pacman) # Load the package in R environment
p_load(epiphy, tidyverse, cowplot, patchwork) # Install and load additional package in R
```


# Data

*citrus_ctc* datasets contains information about the **incidence** of citrus tristeza virus (CTV) disease in three citrus fields.
```{r warning=F}
iv3 <- citrus_ctv$IVI3and4
head(iv3)
tail(iv3)
str(iv3)
```
On these two examples, *x* and *y* correspond to the spatial coordinates of the sampling units distributed in a regular two-dimensional grid. 

 - *x* and *y* are the row id and the within-row sampling unit id, respectively. 
 
 - *t* variable stands for the assessment time or date. 
 
 - *i* variable corresponds to the number of recorded individuals (presence or absence of symptomatics plants) in each sampling unit. 
 
 - *n*, which is only there in incidence data sets, is the total number of individuals in a sampling unit. 

As in the raw CTV incidence data set, *n* = 1,  means that each sampling unit contains only one plant and so *i* can only be equal to 0 (the plant is healthy) or 1 (the plant is diseased).

## handling
```{r warning=F}
inc <- incidence(iv3)
inc
plot(inc, tile = F)

inc4 <- clump(inc, unit_size = c(x = 2, y = 2))
plot(inc4, tile = F, size = 4)
```

```{r warning=F}
inc4_91 <- split(inc4, by = "t")[[1]]
plot(inc4_91)

inc4_93 <- split(inc4, by = "t")[[3]]
plot(inc4_93)
```


# Statistical methods

## Aggregation indices
```{r warning=F}
inc4_91_idx <- agg_index(inc4_91)
chisq.test(inc4_91_idx)
```

```{r warning=F}
inc4_93_idx <- agg_index(inc4_93)
chisq.test(inc4_93_idx)
```

## Distribution fitting
```{r warning=F}
inc4_91_dist <- fit_two_distr(inc4_91)
summary(inc4_91_dist)
```

```{r warning=F}
inc4_93_dist <- fit_two_distr(inc4_93)
summary(inc4_93_dist)
```

```{r warning=F}
dist_91 <- plot(inc4_91_dist)
dist_93 <- plot(inc4_93_dist)
```

## Binary power law

### Incidence dataset
CTV incidence data for three orchards in eastern Spain reported for consecutive years.

```{r warning=F, echo=F}
inc6_spl <- split(inc4, unit_size = c(x = 3, y = 2))
inc_plaw <- power_law(inc6_spl)
coef(summary(inc_plaw))
bpl <- plot(inc_plaw)
```

### Count dataset

**Counts** of 554 aphids of the species *Sitobion avenae*, sampled on 28 June 1996 in a 250 x 180-m field of winter wheat near Wimborne, Dorset, UK. The 63 sampling units, made of the inspection of five tillers each, were located on a 9 x 7 rectangular grid at intervals of 30 m.

```{r warning=F, echo=F}
coun <- count_data(arthropods[arthropods$x <= 6, ])
cou <- split(coun, unit_size = c(x = 3, y = 3))
cou_plaw <- power_law(cou)
coef(summary(cou_plaw))
tpl <- plot(cou_plaw)
```

```{r}
bpl+tpl
```


## Spatial hierarchy
```{r error=F, warning=F}
plot(inc4_91)
plot(threshold(inc4_91))

incall_low <- split(inc4, unit_size = c(x = 3, y = 2))
incall_high <- lapply(incall_low, threshold)
incall_sphier <- spatial_hier(incall_low, incall_high)

plot(incall_sphier)
```

## SADIE
```{r error=F, warning=F, message=F}
iv3_m <- iv3 %>% 
  filter(t == 1991) %>%
  mutate(xm = x*5, ym = y*5)
head(iv3_m)
str(iv3_m)

inc_m <- incidence(iv3_m, mapping(x = xm, y = ym))
plot(inc_m)

res <- sadie(inc_m, nperm = 10, index = "all")

summary(res)
plot(res)
plot(res, isoclines = TRUE)
```

```{r error=F, warning=F}
iv93_m <- iv3 %>% 
  filter(t == 1993) %>%
  mutate(xm = x*5, ym = y*5)
head(iv93_m)
str(iv93_m)

inc_m <- incidence(iv93_m, mapping(x = xm, y = ym))
plot(inc_m)

res93 <- sadie(inc_m, nperm = 10, index = "all")

summary(res93)
plot(res93)
plot(res93, isoclines = TRUE)
```



# Reference
https://chgigot.github.io/epiphy/articles/epiphy.html